Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-09-08 draft.tex
revised results plot, now contains BMI

\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
\affiliation{\sup{1}Department of Simulation and Graphics, University of Magdeburg}
\affiliation{\sup{2} Institut for Community Medicine, University of Greifswald}
\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald}
\email{\{paul, sylvia\}@isg.cs.uni-magdeburg.de, \{katrin.hegenscheid, voelzke\}@uni-greifswald.de}
}

@@

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model is initialized using three user-placed landmarks.
%%
A click on the L3 vertebra center initialized the model, two clicks on the top and bottom of the vertebra describes an initial height estimation.
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
%%
The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
%%
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%% BLIND
%In a previous work (\cite{Klemm2013VMV}), we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
The centerline representation of the lumbar spine canal was extracted from the detection model (Fig.~\ref{fig:centerline}~a) using the approach of \cite{Klemm2013VMV}.
%%
Using the Frenet frame, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Normalized Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{i}$. We refer to total curvature from now on as \emph{curvature}.
%%
	\item \emph{Normalized Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{i}$. We refer to total torsion from now on as \emph{torsion}
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Age Groups}
%%
Gender could be classified for each age group using \emph{mean curvature} and \emph{curvature angle}.
%%
The accuracy varies between $0.35$ (\emph{$43.8-54$ years of age}, 3 decision rules) to $0.27$ (\emph{$21-43.8$ years of age}, 6 decision rules).
%%
The starting age of smoking could be characterized well with an error rate between $0.25$ to $0.3$ for all \emph{age} groups except of \emph{$21.1-43.8$ years of age}.
%%
The result however is probably over fitted to the data due to tree sizes between $16$ and $23$.
%%
\com{Feedback der Greifswalder notwendig, warum das so sein koennte.}
%%

@@

%TODO: Muss das wirklich noch rein?
% \com{Manchmal kann es passieren, dass die Quantile fuer eine Variable gleich sind - z.B. bei Dauer Jahre Schwerarbeit, weil viele Probanden hier eben eine 0 haben. Deswegen kann das Ding nicht in vier Qunatile unterteilt werden, die ersten 3 sind 0. Deswegen wurden nur die einzigartigen Unterscheidungen genommen (statt \emph{quantile(myVar)} wurde \emph{unique(quantile(myVar))} verwendet.)}

\com{Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)}

@@

Another observation during the analysis is very large decision trees achieving low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\subsection{Proposed Features}
\label{sec:ProposedFeatures}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.475\textwidth]{figures/proposed-features}
  \caption{
	Proposed features.
	%%
	(a) dented vertebraes are a sign of heavy stress and bone absorption.
	%%
	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
	%%
	(c) Spine canal thickness is associated with herniated disks.
	%%
	}
  \label{fig:proposedFeatures}
\end{figure}
%%
\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Zitat Frau Hegenscheid fragen}.
%%
We have to capture earlier signs of pathological change in order to derive better predictive models.
%%

@@

\noindent \com{TODO: Marko hinzufuegen} SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.

>>>>

\author{\authorname{Paul Klemm\sup{1}, Sylvia Glaßer\sup{1}, Kai Lawonn\sup{1}, Marko Rak\sup{1} Henry Völzke\sup{2}, Katrin Hegenscheid\sup{3} and Bernhard Preim\sup{1}}
\affiliation{\sup{1}Department of Simulation and Graphics, University of Magdeburg}
\affiliation{\sup{2} Institut for Community Medicine, University of Greifswald}
\affiliation{\sup{3} Institute for Diagnostic Radiology and Neuro-Radiology, University of Greifswald}
\email{\{paul, sylvia\}@isg.cs.uni-magdeburg.de, \{katrin.hegenscheid, voelzke\}@uni-greifswald.de}
}

@@

A hierarchical finite element method was used to detect the lumbar spine in the MRI data \cite{Rak2013}.
%%
The tetrahedron-based finite element model is initialized using three user-placed landmarks.
%%
A click on the L3 vertebra center initialized the model, two clicks on the top and bottom of the vertebra describes an initial height estimation.
%%
The registered model captures information about the lumbar spine canal shape and the position of the L1-L5 vertebrae \cite{Klemm2013VMV} (Fig.~\ref{fig:centerline}~a).
%%
Each subject's model is mapped to a common reference frame using procrustes analysis to compensate for differences in translation, 
rotation and scaling.
%%
The detection fails for several subjects due imprecise initialization, imaging artifacts or strongly deformed spines.
%%
$2,540$ lumbar spine models were obtained and constitute the foundation of this work.
%%
% \com{From VIS'14 Paper}
% The lumbar spine was detected in the image data using a hierarchical finite element method by Rak et al. [32].
% %%
% This semiautomatic method requires the user to initialize the tetrahedron-based finite element models (FEM) with a click on the L3 vertebra.
% %%
% Two user-defined landmarks on the top and bottom of the L3 vertebra describe an initial model height estimation.
% %%
% The model uses a weighted sum of T1 and T2-weighted MR images to detect the lumbar spine shape.
% %%
% Once registered, it captures information about the shape of the lumbar spine canal as well as the position of the L1-L5 vertebrae [21].
% %%
% Due to incorrect initialization, strongly deformed spines, contrast differences and artifacts, the model was not able to detect lumbar spines for all subjects.
% %%
% We obtained and worked with 2,540 tetrahedron models of the lumbar spine.
% %%
% For clustering, we extracted the centerline of the lumbar spine canal, which captures information about lordosis and scoliosis (the medical terms for spine curvature) [21].
% %%

%\com{Marko, kannst du mal bitte schauen, ob das hier so korrekt ist, bzw. ob man das konkreter formulieren kann?}
We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%% BLIND
%In a previous work (\cite{Klemm2013VMV}), we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
The centerline representation of the lumbar spine canal was extracted from the detection model (Fig.~\ref{fig:centerline}~a) using the approach of \cite{Klemm2013VMV}.

%%
Using the Frenet frame, we calculated the following metrics from the model (Fig.~\ref{fig:centerline}~b) \cite{Frenet}:
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Normalized Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{curvature_i}{i}$. We refer to total curvature from now on as \emph{curvature}.
%%
	\item \emph{Normalized Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum_{i=1}^I \frac{torsion_i}{i}$. We refer to total torsion from now on as \emph{torsion}
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Age Groups}
%%
Gender could be classified for each age group using \emph{mean curvature} and \emph{curvature angle}.
%%
The accuracy varies between $0.35$ (\emph{$43.8-54$ years of age}, 3 decision rules) to $0.27$ (\emph{$21-43.8$ years of age}, 6 decision rules).
%%
The starting age of smoking could be characterized well with an error rate between $0.25$ to $0.3$ for all \emph{age} groups except of \emph{$21.1-43.8$ years of age}.
%%
The result however is probably over fitted to the data due to tree sizes between $16$ and $23$.
%%
\com{Warum?}
%%

@@

% \com{Manchmal kann es passieren, dass die Quantile fuer eine Variable gleich sind - z.B. bei Dauer Jahre Schwerarbeit, weil viele Probanden hier eben eine 0 haben. Deswegen kann das Ding nicht in vier Qunatile unterteilt werden, die ersten 3 sind 0. Deswegen wurden nur die einzigartigen Unterscheidungen genommen (statt \emph{quantile(myVar)} wurde \emph{unique(quantile(myVar))} verwendet.)}

%TODO: Muss das wirklich noch rein?
%\com{Auswertung der Anteile der Parameter an den Ergebnissen (Torsion, Curvature, etc. - welches ist der dominanteste Parameter?)}

@@

Another observation during the analysis is very large decision trees achieving low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\subsection{Proposed Features}
\label{sec:ProposedFeatures}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.475\textwidth]{figures/proposed-features}
  \caption{
	Proposed features.
	%%
	(a) dented vertebraes are a sign of heavy stress and bone absorption.
	%%
	(b, c) the spine canal shape can be used to characterize scoliosis (curvature sagittal) and lordosis (curvature coronal) as well as the \emph{Cobb-Angles} \cite{Cobb}.
	%%
	(c) Spine canal thickness is associated with herniated disks.
	%%
	}
  \label{fig:proposedFeatures}
\end{figure}
%%
\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Referenz?}.
%%
We have to capture earlier signs of pathological change in order to derive better predictive models.
%%

@@

\noindent SHIP is part of the Community Medicine Research net of the University of Greifswald, Germany, which is funded by the Federal Ministry of Education and Research (grant no. 03ZIK012), the Ministry of Cultural Affairs as well as the Social Ministry of the Federal State of Mecklenburg-West Pomerania. Whole-body MR imaging was supported by a joint grant from Siemens Healthcare, Erlangen, Germany and the Federal State of Mecklenburg-Vorpommern. The University of Greifswald is a member of the ‘Centre of Knowledge Interchange’ program of the Siemens AG. This work was supported by the DFG Priority Program 1335: Scalable Visual Analytics.
